{\large Gravitational wave sources in our Galactic backyard:}\\{\normalsize Models for detecting BHBH, BHNS and NSNS binaries with LISA

### 2021-11-19 paper/tex/sections/variations.tex
WDWD changes in discussion

In models \modAlphaLowest{}-\modAlphaHighest{} we alter the common-envelope efficiency parameter $\alpha_{\rm CE}$ to $0.1, 0.5, 2.0$ and $10.0$ respectively. Values smaller than 1 may represent cases where not all energy is used efficiently to unbind the envelope, for example when part of the energy escapes in the form of radiation or if part is used to impart additional kinetic energy in the ejecta. Values larger than 1 may represent cases where additional energy sources can be tapped into, such as for example jets powered by accretion. The variations can also be seen as a way to cover uncertainties in estimates for the binding energy itself.

Increasing $\alpha_{\rm CE}$ makes common-envelope ejection more efficient or, in other words, less orbital shrinkage is needed to successfully eject the envelope. This has two consequences. (1) Firstly, a larger fraction of systems avoids merging during the CE phase. This increases the overall number of DCOs. (2) Secondly, the systems that survive the CE phase are wider, possibly too wide to become detectable as gravitational wave sources. So, while the first consequence favours the formation of DCOs, the second consequence disfavours the formation of DCOs that are tight enough to be detected.

These two opposing effect result in a fine tuning situation. Only a very small subset of progenitor systems that have the right orbital parameters prior to the CE phase to successfully produce detectable systems. Changing  $\alpha_{\rm CE}$ moves and changes this window in the parameter space that successfully leads to the formation of detectable systems. How the number of the detectable systems changes depends on whether the relevant parameter space grows or shrinks and on how well the relevant part of the parameters space is populated. Fully unravelling these effects and how they interplay with the assumed star formation history is beyond our scope (and possibly not even of large relevance given the severe simplifications). We will limit the further discussions to simply stating the trends we observe.

@@

In agreement with other studies, we find that this treatment leads to a large increase formation rate of BHBHs and with it the number of systems that LISA can detect. This is because BHBHs originate from the most massive progenitors, which expand significantly during the Hertzsprung gap phase in our simulations. The progenitors likely initiate interaction during this phase. Our predictions for the number of detectable systems increases by a factor 2. The less massive progenitors of BHNSs are less strongly affected, and for NSNS we do not see a significant increase.

\paragraph{Case BB mass transfer} For completeness and consistency with \citet{Broekgaarden+2021}, we have also included model variations that consider uncertainties in case BB mass transfer. This is a phase of mass transfer where the donor star has already lost its hydrogen envelope in a prior interaction, but fills its Roche lobe again as it expands during helium shell burning phase \citep[e.g.][]{Dewi+2003, Vigna-Gomez+2018}. This is of particular interest for the formation of NSs, as their lower mass progenitors are swelling the most during this phase \citep[e.g.][and references therein]{Laplace+2020}. Population synthesis studies find that nearly all NSNS systems might from through a phase of case BB mass transfer \citep{Vigna-Gomez+2018}.

In model \modCaseBB{} we effectively exclude any systems that evolve through case BB mass transfer. As expected, this kills nearly all NSNS systems. We also find a heavy reduction of the BHNS systems by a factor 6. The BHBH systems are not affected, as expected, since case BB mass transfer does not play a role for high mass progenitors.

Finally, in model \modCaseBBOpt{}, we consider a situation where case BB mass transfer always leads to unstable mass transfer and initiates a CE phase in combination with the optimistic treatment for CE. This model leads to the highest predictions for the detections among all variations that we have considered.

@@

\subsubsection{Stellar winds}
Mass loss in the form of stellar winds or eruptions is also a main uncertainty. It affects by how much stars can grow in size and it affects the final core masses. \citet{Broekgaarden+2021} provides us with model variations that consider variations in the mass loss by naked helium (Wolf-Rayet like) stars. We find that a reduction of the wind mass loss has very little effect on our predictions. This is the consequence of several effects that cancel each other.

Firstly, the decreased Wolf-Rayet like winds mean that the DCOs (particularly those containing BHs) are generally more massive and so more detectable in LISA. Secondly, one may expect that LISA sources in model \modWRLow{} would be higher frequency than in our fiducial model as decreased winds generally result in tighter binaries. However, though this is the case at DCO formation, we find that by the time the sources have evolved until they are observable by the LISA mission, they have lower orbital frequencies than in our fiducial model. This is because the reduced winds allow DCOs to be formed at higher metallicity and, therefore, at more recent times. This means that the DCOs do not evolve for as long before the LISA mission and so remain at lower frequencies (wider separations) thus making them less detectable.

@@

In model \modWRHigh{} we instead \textit{increase} the efficiency of Wolf-Rayet winds by a factor of 5. We note that though there is currently no observational support for such a strong increase in the stellar winds of Wolf-Rayet stars, it is nevertheless still interesting to consider. In this model the detection rate of BHBHs decreases by over a factor of 10 and BHNSs by over a factor of 2 whilst the NSNS rate is relatively unchanged. Increasing the efficiency of WR winds strongly decreases the final masses of DCOs. This means that many progenitors that would have formed LISA sources under our fiducial assumptions would not have enough mass to produce a DCO, or produce a NS instead of a BH (therefore this effect is stronger on BHBHs than NSNSs). Moreover, DCOs that are formed tend to be less massive and therefore less detectable.

@@

Alternatively, the blue curve in Fig.~\ref{fig:lower_mass_gap_variation} shows the same distribution but for the \textit{rapid} remnant mass prescription \citep{Fryer+2012}, which we use in model \modRapid{}. In this case, no compact objects are formed (and therefore, detected) in the lower mass gap. From the stark difference between these models, it is clear that it is difficult at this point to say with any certainty what fraction of systems LISA will detect in the lower mass gap given the highly uncertain formation rate of systems in this mass range.
%
However, it is important to highlight that \textit{if} DCOs are formed with components in the lower mass gap, LISA \textit{will} be able to detect them. And thus, LISA could be a useful instrument for providing evidence for the existence or non-existence of a lower mass gap based on the mass distribution of detected DCOs.

@@

In Fig.~\ref{fig:wr_wind_mass_variations} we show the effect that changing the efficiency of Wolf-Rayet winds has on the individual component mass distribution. Decreasing the Wolf-Rayet wind efficiency allows the formation of more massive DCOs in the Milky Way and, indeed, we see that the distribution extends to $25 \unit{M_{\odot}}$ and fewer detectable systems are formed at low masses. By contrast, increasing the Wolf-Rayet efficiency by a factor of 5 strongly disfavours the formation of systems at high masses and approximately 85\% of detectable systems have masses below $5 \unit{M_{\odot}}$. These three distributions are very distinct and so it is possible that the mass distribution of LISA could help to constrain the efficiency of Wolf-Rayet winds.

>>>>

In models \modAlphaLowest{}-\modAlphaHighest{} we alter the common-envelope efficiency parameter $\alpha_{\rm CE}$ to $0.1, 0.5, 2.0$ and $10.0$ respectively. Values smaller than 1 may represent cases where not all energy is used efficiently to unbind the envelope, for example when part of the energy escapes in the form of radiation or if part is used to impart additional kinetic energy in the ejecta \citep[e.g.\ ][]{Ivanova+2013, Nandez+2016}. Values larger than 1 may represent cases where additional energy sources can be tapped into, such as for example jets powered by accretion \citep[e.g.\,][]{Schreier+2021}. The variations can also be seen as a way to cover uncertainties in estimates for the binding energy itself.

Increasing $\alpha_{\rm CE}$ makes common-envelope ejection more efficient or, in other words, less orbital shrinkage is needed to successfully eject the envelope. This has two consequences. (1) A larger fraction of systems avoids merging during the CE phase. This increases the overall number of DCOs. (2) The systems that survive the CE phase are wider, possibly too wide to become detectable as gravitational wave sources \citep[e.g.\,][]{Klencki+2021}. So, while the first consequence favours the formation of DCOs, the second consequence disfavours the formation of DCOs that are tight enough to be detected.

These two opposing effect result in a fine tuning situation. Only a very small subset of progenitor systems have the right orbital parameters prior to the CE phase to successfully produce detectable systems. Changing $\alpha_{\rm CE}$ moves and changes this window in the parameter space that successfully leads to the formation of detectable systems. How the number of the detectable systems changes depends on whether the relevant parameter space grows or shrinks and on how well the relevant part of the parameters space is populated. Fully unravelling these effects and how they interplay with the assumed star formation history is beyond our scope (and possibly not even of large relevance given the severe simplifications). We will limit the further discussions to simply stating the trends we observe.

@@

In agreement with other studies, we find that this treatment leads to a significant increase in the formation rate of BHBHs, by a factor of two. This is because the progenitors expand significantly during the Hertzsprung gap phase in our simulations. In our fiducial simulation, all progenitors that initiate unstable interaction during this phase would end as stellar mergers, while in this variation they will survive. The progenitors of BHNSs and NSNSs are less strongly affected with an increase about 30\% increase.

\paragraph{Case BB mass transfer} In models \modCaseBB{} and \modCaseBBOpt{} we consider uncertainties in case BB mass transfer. This is a phase of mass transfer where the donor star has already lost its hydrogen envelope in a prior interaction, but fills its Roche lobe again as it expands during helium shell burning phase \citep[e.g.][]{Dewi+2003, Tauris+2015, Tauris+2017}. This is of particular interest for the formation of NSs, as their lower mass progenitors are swelling the most during this phase \citep[e.g.][and references therein]{Laplace+2020}. Population synthesis studies find that nearly all NSNS systems form through a phase of case BB mass transfer \citep{Vigna-Gomez+2018}.

In model \modCaseBB{} we enforce that case BB mass transfer is always unstable, such that it always leads to a CE. This effectively excludes any systems that evolve through case BB mass transfer since all common-envelopes initiated in this way have donors in the Hertzsprung gap and thus end as stellar mergers under the pessimistic CE assumption. Therefore, this kills nearly all NSNS systems. We also find a heavy reduction of the BHNS systems by a factor 6. The BHBH systems are not affected, as expected, since case BB mass transfer does not play a role for high mass progenitors.

Finally, in model \modCaseBBOpt{}, we again enforce that case BB mass transfer is always unstable, but in combination with the optimistic treatment for CE (essentially combining models \modCaseBB{} and \modOpt{}). This allows the systems formed through case BB (as well as any other systems that have HG donors for common-envelopes) to survive the CE phase. We find that this model leads to the highest predictions for the detections among all variations that we have considered.

@@

\subsubsection{Stellar winds}
Mass loss in the form of stellar winds or eruptions is also a main uncertainty. It affects by how much stars can grow in size and it affects the final core masses. We consider variations in the mass loss by naked helium (Wolf-Rayet like) stars and choose to vary the efficiency of these winds between $0.1$ and $5$, to account for uncertainties in the derived rates \citep[e.g.][]{Vink+2017, Shenar+2019, Hamann+2019, Sander+2020}.

We find that a reduction of the wind mass loss has very little effect on our predictions. This is the consequence of several effects that cancel each other. Firstly, the decreased Wolf-Rayet like winds mean that the DCOs (particularly those containing BHs) are generally more massive and so more detectable in LISA. Secondly, one may expect that LISA sources in model \modWRLow{} would be higher frequency than in our fiducial model as decreased winds generally result in tighter binaries. However, though this is the case at DCO formation, we find that by the time the sources have evolved until they are observable by the LISA mission, they have lower orbital frequencies than in our fiducial model. This is because the reduced winds allow DCOs to be formed at higher metallicity and, therefore, at more recent times. This means that most DCOs do not evolve for as long before the LISA mission and so remain at lower frequencies (wider separations) thus making them less detectable.

@@

In model \modWRHigh{} we instead \textit{increase} the efficiency of Wolf-Rayet winds by a factor of 5. In this model the detection rate of BHBHs decreases by over a factor of 10 and BHNSs by over a factor of 2 whilst the NSNS rate is relatively unchanged. Increasing the efficiency of WR winds widens the orbit and decreases the final masses of DCOs. This means that some progenitors that would have formed LISA sources under our fiducial assumptions would not have enough mass to produce a DCO, or produce a NS instead of a BH. The effect is strongest for the progenitors of black holes, since these are most strongly affected by Wolf-Rayet winds. This effect is less pronounced in NSs since the rate of mass loss for the progenitors of these systems is low enough that changing by a factor of 5 still does not impact the final fate. Moreover, DCOs that are formed tend to be less massive and therefore less detectable.

@@

Alternatively, the blue curve in Fig.~\ref{fig:lower_mass_gap_variation} shows the same distribution but for the \textit{rapid} remnant mass prescription \citep{Fryer+2012}, which we use in model \modRapid{}. In this case, no compact objects are formed (and therefore, detected) in the lower mass gap. From the stark difference between these models, it is clear that it is difficult at this point to say with any certainty what fraction of systems LISA will detect in the lower mass gap given the highly uncertain formation rate of systems in this mass range.

However, it is important to highlight that \textit{if} DCOs are formed with components in the lower mass gap, LISA \textit{will} be able to detect them. And thus, LISA could be a useful instrument for providing evidence for the existence or non-existence of a lower mass gap based on the mass distribution of detected DCOs.

@@

In Fig.~\ref{fig:wr_wind_mass_variations} we show the effect that changing the efficiency of Wolf-Rayet winds has on the individual component mass distribution. Decreasing the Wolf-Rayet wind efficiency allows the formation of more massive DCOs in the Milky Way and, indeed, we see that the distribution extends to $25 \unit{M_{\odot}}$ and relatively fewer detectable systems are formed at low masses. By contrast, increasing the Wolf-Rayet efficiency by a factor of 5 strongly disfavours the formation of systems at high masses and approximately 85\% of detectable systems have masses below $5 \unit{M_{\odot}}$. These three distributions are very distinct and so it is possible that the mass distribution of LISA could help to constrain the efficiency of Wolf-Rayet winds.
